---
title: "Store Sales - Time Series Forecasting"
author: "NJUGUNA J.M"
date: "2024-10-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# Load required libraries
library(data.table)
library(dplyr)
library(ggplot2)
library(forecast)
library(caret)
library(zoo)
library(lubridate)
```

INTRODUCTION:

In this project, we aim to predict sales for thousands of products in Favorita grocery stores using time-series forecasting with machine learning.

Loading and Exploring the Data:
First let's load the dataset

```{r}
# Load the train and test datasets
train <- read_csv("train.csv")
test <- read_csv("test.csv")

# Load supplementary data
stores <- read_csv("stores.csv")
oil <- read_csv("oil.csv")
holidays_events <- read_csv("holidays_events.csv")

# check the structure of the dataset
str(train)
str(test)
str(holidays)
str(oil)
str(stores)
```

Data Preprocessing:
We will perform feature engineering, handle missing values, and create lag features for the dataset.

   Feature Engineering (Including lagged sales and holiday indicators)
```{r}
# Convert date column to Date type
train$date <- as.Date(train$date)
test$date <- as.Date(test$date)

# Add year, month, and weekday features
train$year <- year(train$date)
train$month <- month(train$date)
train$day <- day(train$date)
train$weekday <- wday(train$date)

test$year <- year(test$date)
test$month <- month(test$date)
test$day <- day(test$date)
test$weekday <- wday(test$date)

# Merge holidays and oil data
train <- merge(train, holidays, by = "date", all.x = TRUE)
train <- merge(train, oil, by = "date", all.x = TRUE)
test <- merge(test, holidays, by = "date", all.x = TRUE)
test <- merge(test, oil, by = "date", all.x = TRUE)

# Create a holiday indicator (1 = holiday, 0 = no event)
train$holiday_indicator <- ifelse(is.na(train$holiday) | train$holiday == "No Event", 0, 1)
test$holiday_indicator <- ifelse(is.na(test$holiday) | test$holiday == "No Event", 0, 1)

# Create lag features (example: 1-day lag)
train <- train %>%
  group_by(store_nbr, family) %>%
  arrange(date) %>%
  mutate(lag_1 = lag(sales, 1), lag_7 = lag(sales, 7), lag_30 = lag(sales, 30) )

# To handle later in the model fitting
test$lag_7 <- NA  
test$lag_1 <- NA 
test$lag_30 <- NA
```
  Handle missing values
```{r}
# Use na.locf() to forward-fill missing lag values in train data
train$lag_1 <- na.locf(train$lag_1, na.rm = FALSE)
train$lag_7 <- na.locf(train$lag_7, na.rm = FALSE)
train$lag_30 <- na.locf(train$lag_30, na.rm = FALSE)

# In test data, we will create lag features during prediction.
```

Model Training with ARIMA:
We will now train an ARIMA model for each store_nbr and family. We’ll use the auto.arima() function to automatically select the best ARIMA model for each time-series.

  Select Time-Series Data:
We need to group the data by store_nbr and family to train a separate ARIMA model for each product family in each store.
```{r}

# Grouping data by store and product family for ARIMA modeling
store_family_groups <- unique(train[, .(store_nbr, family)])

# Create a list to store ARIMA models for each group
arima_models <- list()
```

  Train ARIMA Models:
We will loop over each store_nbr and family and fit an ARIMA model on the sales data.

```{r}
for (i in 1:nrow(store_family_groups)) {
  store_nbr <- store_family_groups$store_nbr[i]
  family <- store_family_groups$family[i]
  
  # Filter data for this group
  train_subset <- train %>% filter(store_nbr == store_nbr, family == family)
  
  # Ensure data is ordered by date
  train_subset <- train_subset %>% arrange(date)
  
  # Create a time-series object
  ts_data <- ts(train_subset$sales, frequency = 365, start = c(min(train_subset$year), min(train_subset$month)))
  
  # Fit ARIMA model using auto.arima()
  arima_model <- auto.arima(ts_data)
  
  # Store the ARIMA model
  arima_models[[paste(store_nbr, family, sep = "_")]] <- arima_model
}
```
 Examine a Fitted Model:
We can examine one of the fitted ARIMA models for a specific store and family.

```{r}
# check ARIMA model for store 1, family 'Dairy'
arima_models[["1_Dairy"]]
```
 Model Evaluation:
Evaluate the model using time-series cross-validation

```{r}
# forecasting sales for the next 15 days using one ARIMA model (store 1, Dairy family)
store_nbr <- 1
family <- "Dairy"
arima_model <- arima_models[[paste(store_nbr, family, sep = "_")]]
forecasted_sales <- forecast(arima_model, h = 15)
plot(forecasted_sales)
```
 Forecasting on Test Data:
Use the fitted ARIMA models to predict sales for each store_nbr and family in the test dataset.

```{r}
# Initialize a dataframe to store test predictions
test$predicted_sales <- NA

for (i in 1:nrow(store_family_groups)) {
  store_nbr <- store_family_groups$store_nbr[i]
  family <- store_family_groups$family[i]
  
  # Filter test data for this group
  test_subset <- test %>% filter(store_nbr == store_nbr, family == family)
  
  # Get the fitted ARIMA model
  arima_model <- arima_models[[paste(store_nbr, family, sep = "_")]]
  
  # Forecast sales for the next 15 days
  forecasted_sales <- forecast(arima_model, h = nrow(test_subset))
  
  # Assign forecasted sales to the test data
  test$predicted_sales[test$store_nbr == store_nbr & test$family == family] <- forecasted_sales$mean
  
}
```

 Conclusion:
In this project, we used ARIMA to predict sales for products across various Favorita grocery stores in Ecuador. We explored time-series data and successfully trained ARIMA models for accurate forecasting.
```{r}

### Key Steps Included:
- **Data Preprocessing**: Feature engineering, handling missing values, and preparing the data.
- **ARIMA Model Training**: Fit an ARIMA model for product family in each store using `auto.arima()`.
- **Forecasting**: Use the fitted ARIMA models to forecast sales for the test set.






